Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-03 joint_inf.tex
address all comments and update responses

The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry} or a relative likelihood \citep{Goldman1990-dk}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}_\nCols$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}_\nCols, \hat{t}_\nCols) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
%EM ancestral state splits not introduced-- can we just say "ancestral states"? Note that splits appear below as well.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the conditional likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.

@@

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.

@@

%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
\caption{
    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:incons-analytic}
\end{figure}

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we estimate the interior branch length to be zero.

@@

%EM This is an important paragraph but the meat is hidden in the middle.
%EM We need to start with a topic sentence that says that For branch lengh, we find a systematic bias even in the short-branch case (Figure ref).
%EM Then we can talk about details such as discontinuities and how marginal works great.
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
Because of these discontinuities, an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
%EM I suggest compacting the next few sentences.
The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
%EM What is the "general case" here?
Similarly, for the general case in estimating branch lengths, we are only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

>>>>

The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry} or a relative likelihood \citep{Goldman1990-dk}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}_\nCols$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}_\nCols, \hat{t}_\nCols) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
%EM ancestral state splits not introduced-- can we just say "ancestral states"? Note that splits appear below as well.
%das: I removed "splits" where I could, but kept it right before the theorem. Thoughts?
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral states (Table~\ref{tab:farris_likelihoods}) to obtain each conditional likelihood term, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.

@@

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits---the explicit definition for $\boldsymbol\xi$ is given in the Appendix.
Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.

@@

%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
%das: good idea; I think this will also appease reviewer 1.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states.
This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we estimate the interior branch length to be zero.

@@

%EM This is an important paragraph but the meat is hidden in the middle.
%EM We need to start with a topic sentence that says that For branch lengh, we find a systematic bias even in the short-branch case (Figure ref).
%EM Then we can talk about details such as discontinuities and how marginal works great.
%das: okay, reworked.
In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
The estimates for $\hat{p}_w$ in this case look reasonable at first, compared with those estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), though joint inference predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
Inference on the integrated likelihood performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
Moreover, the errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
%das: yeah, it was kind of hand-wavy.
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
%EM I suggest compacting the next few sentences.
The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
%EM What is the "general case" here?
%das: did some scalpeling and rewording
For the case of estimating branch lengths, we are able to provide empirical estimates for regions of overestimation that are tighter than the analytical bounds, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
